### 2017-11-29 manuscript.rmd
modified discussion

### Species phylogeny and hypothesis testing
Specimens were collected in the eastern Pacific Ocean, Mediterranean, and the Gulf of California (table 1). All sequence data have been deposited in the sequence read archive (SRA). The analyses presented here consider 34 siphonophore species and 9 outgroup species. This includes new data for 30 species. Summary statistics for expression libaries are given in supplementary table x. In the final analyses, we sampled 1,071 genes to generate a supermatrix with 60% occupancy and a length of 378,468 amino acids (gene occupancy matrix - supplementary figure 1). 
```{r specimen_summary, results='asis', echo=FALSE, comment=NA, warning=FALSE}
specimen_data <- read.csv("supplementary_info/Specimen_data.csv")
names(specimen_data)<-c("New data","Species","ID","SRA Number", "Library ID", "Library type","Tissue","Sequencer", "Sequencing center", "Sample prep","Collected by","Collection date","Specimen id","Depth (m)","Geo loc name","Lat Lon","Voucher Type","Voucher location","Voucher Cat No.","Collection mode")

@@

```
These findings are largely consistent with a previous analysis based on two genes (16S and 18S ribosomal RNA) (figure 3) [@Dunn:2005dy]. There is strong support for the Cystonectae as sister to all other siphonophores, the Codonophora. We also find strong support for Calycophorae nested within the paraphyletic "Physonectae", and for the Apolemidae as sister to all other Codonophora. Multiple nodes that were not resolved in the previous two-gene analysis do recieve strong support in this 1,071-gene transcriptome analysis. There is strong support for the Pyrostephidae as sister to all non-Apolemid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, the Calycophorae and a clade we call the Euphysonectae, which includes the remaining non-Apolemid, non-Pyrostephid historically recognized Physonectae. Given the relatively shallow sampling of this analysis, we define the Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to the Calycophorae.

In this phylogeny, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in previous phylogeny) are sister to the Agalmatidae *sensu stricto*, a clade restricted to *Agalma*, the Athorybiidae,*Halistemma* and *Nanomia* [@Dunn:2005dy]. In the rDNA study, *Physophora hydrostatica* was sister to the Forskaliidae with low support. *Cordagalma cordiforme* (= *Cordagalma ordinatum* [@pugh2016synopsis]) was previously unresolved, while in this analysis *Cordagalma* sp is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae *sensu stricto*. *Cordagalma* has traditionally been placed within the Agalmatidae *sensu lato* [@totton1965synopsis], although the Agalmatidae is considered to be a catch-all family in need of revision [@pugh2006taxonomic,@pugh1998re, @Dunn:2005dy]. Within the calycophorans, taxon sampling is shallower, however there is broad agreement with the previous analysis. *Craseoa lathetica* and *Desmophyes haematogaster* are sister to *Hippopodius hippopus*

@@

@@

The strong phylogenetic signal in the characters traditinally used for taxonomic diagnostics is a positive indicator of the applicability and unambiguity of these characters.

[Citation needed] noticed for the first time that some siphonophores were monoeicous and others were dioiecious. Our analyses in 2005 reconstructed this character and found a great amount of phylogenetic conservatism, with an unambiguous resolution of the MRCA (most recent common ancestor) as dioecious, and the appearance of monoecy in several taxa and clades (including Calycophorae) within the polytomy. Figure 5a shows the evolution of sex distribution in siphonophores under the current better-resolved tree model, and it strongly indicates that monoecy in siphonophores from a dioecious ancestor occured twice, in the branch leading to Calycophorae and in the branch leading to Agalmatids (sensu lato). There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of Calycophorae, with a subsequent derived shift back to dioecy in the *Marrus-Erenna* clade.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types. Other colonial cnidarians (such as Hydractinia) and some bryozoans (example) have been found to have up to X different zooid types [Citation1 , Citation 2]. The siphonophore genus Forskalia has 6 basic zooid types (pneumatophore, nectophore, gastrozooid, palpon, bract, and gonophore), and a total of 10 counting subtypes (4 types of bract, male & female gonophores). Diphyomorphs have more than 1 type of nectophore, while Cystonects have none. Here we reconstruct the evolutionary origins ofthe different zooid types on the present transcriptome tree.

Nectophores are retained modified medusae that Codonophora use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone. Unlike the siphosomal growth zone, the nectosome does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). If fact, with the exception of  *Physalia physalis* (which grows small nectophores near the gonodendra), siphonophore nectophores are exclusively found on the nectosome. It is possible that the MRCA of siphonophores had a nectosome, which has lost on the branch leading to Cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to the Codonophora. The nectosome probably arose as a duplication of the siphosome, followed by functional specialization in propelling the colony. The nectosome has been lost within Codonophora in the genus *Athorybia*.

Following the colony development orientation framework presented in (citation), the nectosome can be located in a dorsal or a ventral position. Our ancestral reconstructions for this character (Supp figure) show that a ventrally-oriented nectosome was the ancestral form in siphonophores, and that a dorsal nectosome has evolved twice independently, in the braches leading to the Agalmatidae (sensu stricto) and the branch leading to the *Bargmannia* species.

All Codonophora (with the exception of *Athorybia* species) have a nectosome, but the number and subtypes of nectophores present varies greatly between species. As shown in Figure X, most Codonophora presents the ancestral nectosome with multiple nectophores of the same subtype. However, Calycophorans evolved a different system with just 2 nectophores of one type. This shift may be associated with the loss of the pneumatophore. Not all Calycophorans remained with this arrangement. The Hippopodidae returned to bearing multiple identical nectophores, many of which are inactive and serve functions of defense (like a shell to retract in) and buoyancy. As in the rest of Calycophorans, the Hippopodids only use 2 nectophores to propel the colony. Another interesting shift occurs in the branch leading to Diphyomorpha, where the 2 nectophores specialize into 2 subtypes, associated with a shift into a vertically aligned position and pointed bell shapes. The 2 types function together in a coupled hydrodynamic system that allows very fast escape responses (Mackie 1964).

Bracts are highly reduced zooids unique to siphonophores, but they are only present in the Codonophora. As with the nectosome, we have ambiguity determining whether the MRCA of siphonophores had bracts or not. The MRCA of Codonophora ahd only one bract subtype, which was lost in Hippopodidae, *Physophora hydrostatica* (however, they are present in its sister species, *P. gilmeri* included in the present phylogeny), and in *Gymnopraia lapislazula*. Bracts are functional for protection of the delicate zooids and to help maintain neutral buoyancy. Some calycophorans are able to actively exclude suphate ions in their bracts to adjust their buoyancy along the colony (Bidigare & Biggs, 1980).

The ancestral siphonophore certainly had a pneumatophore, since both Cystonects and most Codonophorans possess one (Figure 4). This unique zooid fills itself with gas, which helps the colony float and maintain its orientation in the water column. Recent evidence of neural arrangement in the pneumatophore of *Nanomia bijuga*  (Church, 2013) suggests it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis. Despite its multiple biological functions, it was lost in the Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy.

Palpons are modified mouthless gastrozooids used for digestion and circulation of the gastrovascular fluid. They were present in the MRCA of siphonophores (Figure 5b), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in Calycophorans, and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla. Most hydrozoans, including the clade that contains siphonophores, bear simple tentacles (tentacles with no side branches). It is still an open question whether the MRCA of Siphonophora had simple or branched tentacles. The only siphonophores genera regarded as lacking tentilla are *Physalia physalis* and *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstrauct the evolution of this character on the current phylogeny, we find that 70% of simulations support an MRCA bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia*. However, this leaves a 30% support for a simple-tentacled MRCA followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and *{Bargmannia, Diphyes}*.

A key issue here is how we code for absence of tentilla, especially for the case of *Physalia physalis*. The tentacles of this species, when uncoiled, show very prominent, evenly spaced, bulging buttons which contain on their ectoderm all active and functionally arranged nematocysts used by the organism for prey capture. Siphonophore tentilla are complete diverticular brachings of the tentacle ectoderm, mesoglea, and gastrovascular canal (lined by endoderm). Hessinger & Ford 1988 (in the Biology of Nematocysts) described *Physalia*'s buttons as enclosing individual fluid-filled chambers connected by narrow channels to the tentacular canal, lined by endoderm. This suggests they are not just ectodermal swellings, but probably are reduced tentilla. When we code *Physalia physalis* as tentilla bearing, the results for the character reconstraction lead to a more robust support for a tentilla-bearing MRCA followed by a single loss of tentilla in the branch leading to *Apolemiidae* (Supp Figure).

Siphonophore tentilla present an astounding diversity of sizes, shapes, colors, and nematocyst complements, and some have been observed to rapidly uncoil in contact with prey. Future research should explore the evolutionary history of these unique structures.

>>>>

### Species phylogeny and hypothesis testing
Specimens were collected in the eastern Pacific Ocean, Mediterranean, and the Gulf of California (table 1). All sequence data have been deposited in the sequence read archive (SRA). The analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. Summary statistics for expression libaries are given in supplementary table x. In the final analyses, we sampled 1,071 genes to generate a supermatrix with 60% occupancy and a length of 378,468 amino acids (gene occupancy matrix - supplementary figure 1). 
```{r specimen_summary, results='asis', echo=FALSE, comment=NA, warning=FALSE}
specimen_data <- read.csv("supplementary_info/Specimen_data.csv")
names(specimen_data)<-c("New data","Species","ID","SRA Number", "Library ID", "Library type","Tissue","Sequencer", "Sequencing center", "Sample prep","Collected by","Collection date","Specimen id","Depth (m)","Geo loc name","Lat Lon","Voucher Type","Voucher location","Voucher Cat No.","Collection mode")

@@

```
These findings are largely consistent with a previous analysis based on two genes (16S and 18S ribosomal RNA) (figure 3) [@Dunn:2005dy]. There is strong support for the Cystonectae as sister to all other siphonophores, the Codonophora. We also find strong support for Calycophorae nested within the paraphyletic "Physonectae", and for the Apolemidae as sister to all other Codonophora. Multiple nodes that were not resolved in the previous two-gene analysis do recieve strong support in this 1,071-gene transcriptome analysis. There is strong support for the Pyrostephidae as sister to all non-Apolemid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, the Calycophorae and a clade we call the Euphysonectae, which includes the remaining non-Apolemid, non-Pyrostephid "Physonectae". Given the relatively shallow sampling of this analysis, we define the Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to the Calycophorae.

In this phylogeny, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in previous phylogeny) are sister to the Agalmatidae *sensu stricto*, a clade restricted to *Agalma*, the Athorybiidae,*Halistemma* and *Nanomia* [@Dunn:2005dy]. In the rDNA study, *Physophora hydrostatica* was sister to the Forskaliidae with low support. *Cordagalma cordiforme* (= *Cordagalma ordinatum* [@pugh2016synopsis]) was previously unresolved, while in this analysis *Cordagalma* sp is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae *sensu stricto*. *Cordagalma* has traditionally been placed within the Agalmatidae *sensu lato* [@totton1965synopsis], although the Agalmatidae is considered to be a catch-all family in need of revision [@pugh2006taxonomic,@pugh1998re, @Dunn:2005dy].

Within the calycophorans, taxon sampling is shallower, however there is broad agreement with the previous analysis. Calycophorans have in the past been split into convenient groupings of prayomorph and diphyomorph based on morphology (after [@mackie1988siphonophore]). *Craseoa lathetica* and *Desmophyes haematogaster* are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *Craseoa lathetica* and the clade including *Hippopodius hippopus* was unresolved. As in the previous study, this phylogeny indicates that the prayomorphs are paraphyletic.

@@

Siphonophores colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora; however the lack of resolution of deep relationships within the codonophora made it impossible to rule out alternative evolutionary scenarios [@Dunn:2005dy]. The current better-resolved tree model (Figure 5a), indicates that monoecy in siphonophores from a dioecious ancestor occured twice, in the branch leading to Calycophorae and in the branch leading to Agalmatids *sensu lato*. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade.

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969;@cartwright2010character]. For example, the siphonophore genus Forskalia has 6 basic zooid types (pneumatophore, nectophore, gastrozooid, palpon, bract, and gonophore), and a total of 10 counting subtypes (4 types of bract, male & female gonophores). Diphyomorphs have more than 1 type of nectophore, while Cystonects have none. Here we reconstruct the evolutionary origins of the different zooid types on the present transcriptome tree (fig 4).

Nectophores are retained modified medusae that Codonophora use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone (figure 1B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). Siphonophore nectophores are exclusively found on the nectosome -with the exception of  *Physalia physalis* (which has no nectosome, and grows small nectophores near the gonodendra). It is possible that the common ancestor of siphonophores had a nectosome, which has lost on the branch leading to Cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to the Codonophora. It is suggested that the nectosome arose as a duplication of the siphosome, followed by functional specialization in propelling the colony [@dunn2006evolution]. The nectosome has been lost within Codonophora in the genus *Athorybia*.

@@

Following the colony development orientation framework [@haddock_dunn_pugh_2005], the nectosome can be located in a dorsal or a ventral position. Our ancestral reconstructions for this character (S fig 4) show that a ventrally-oriented nectosome was the ancestral form in siphonophores, and that a dorsal nectosome has evolved twice independently, in the braches leading to the Agalmatidae *sensu stricto* and the branch leading to the *Bargmannia* species.

Bracts are highly reduced zooids unique to siphonophores, but they are only present in the Codonophora. As with the nectosome, we have ambiguity determining whether the MRCA of siphonophores had bracts or not. The MRCA of Codonophora had only one bract subtype, which was lost in Hippopodidae, *Physophora hydrostatica* (however, they are present in its sister species, *P. gilmeri* included in the present phylogeny), and in *Gymnopraia lapislazula* (not included in present phylogeny). Bracts are functional for protection of the delicate zooids and to help maintain neutral buoyancy. Some calycophorans are able to actively exclude suphate ions in their bracts to adjust their buoyancy along the colony (Bidigare & Biggs, 1980).

The ancestral siphonophore had a pneumatophore (fig 1B), since both Cystonects and most Codonophorans possess one (fig 4). This unique zooid fills itself with gas, which helps the colony float and maintain its orientation in the water column. Recent evidence of neural arrangement in the pneumatophore of *Nanomia bijuga*  suggests it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The pneumatophore was lost in the Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy.

Palpons are modified mouthless gastrozooids used for digestion and circulation of the gastrovascular fluid [cite]. They were present in the MRCA of siphonophores (Figure 5b), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in Calycophorans, and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla [fig 1A]. Most hydrozoans, including the clade that contains siphonophores, bear simple tentacles (tentacles with no side branches). It is still an open question whether the MRCA of Siphonophora had simple or branched tentacles. The only siphonophores genera regarded as lacking tentilla are *Physalia physalis* and *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an MRCA bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (S Fig 2). However, this leaves a 30% support for a simple-tentacled MRCA followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and *(Bargmannia, Diphyes)*.

A key issue here is how we code for absence of tentilla, especially for the case of *Physalia physalis*. The tentacles of this species, when uncoiled, show very prominent, evenly spaced, bulging buttons which contain on their ectoderm all active and functionally arranged nematocysts used by the organism for prey capture. Siphonophore tentilla are complete diverticular brachings of the tentacle ectoderm, mesoglea, and gastrovascular canal (lined by endoderm). Hessinger & Ford 1988 (in the Biology of Nematocysts) described *Physalia*'s buttons as enclosing individual fluid-filled chambers connected by narrow channels to the tentacular canal, lined by endoderm. This suggests they are not just ectodermal swellings, but probably are reduced tentilla. When we code *Physalia physalis* as tentilla bearing, the results for the character reconstraction lead to a more robust support for a tentilla-bearing MRCA followed by a single loss of tentilla in the branch leading to *Apolemiidae* (S Fig 3).
